# Modules {#modules}

## Environmental drivers {#drmodule}

```{r drList}
dr <- read.csv(here::here("data","cea_modules","drivers_list.csv")) |>
      dplyr::select(-period) |>
      dplyr::distinct() 
nCl <- sum(dr$group == "Climate")
nCo <- sum(dr$group == "Coastal")
nFs <- sum(dr$group == "Fisheries")
nMt <- sum(dr$group == "Marine traffic")
```

We characterized the distribution and intensity of `r nrow(dr)` environmental drivers on the Scotian Shelf during two distinct periods, *i.e.* 2010 to 2015 and 2016 to 2021. Temporal data was unavailable for certain drivers; in such instances, we used the same data for both periods (*e.g.* shipping). Drivers are divided in 4 groups: [climate](drclimate) (n = `r nCl`), [land-based / coastal](drcoastal) (n = `r nCo`), [fisheries](drfisheries) (n = `r nFs`) and [marine traffic](drmarinetraffic) (n = `r nMt`; Table \@ref(tab:drnm)). Drivers were log-transformed to avoid underestimating intermediate intensity values. All drivers were also scaled between 0 and 1 to obtain relative intensities and allow driver comparisons. Scaling was performed using the 99th quantile of intensity distribution as the upper bound to control for extreme values. For more details, see @beauchesne2020 and @halpern2019. The following subsections describes the approach used to evaluate the distribution and the intensity of each driver considered for the cumulative effects assessment. 

```{r drnm}
dplyr::select(dr, group, description) |>
dplyr::arrange(group, description) |>
  knitr::kable(
    col.names = c("Groups", "Drivers"),
    row.names = FALSE,
    caption = "List of environmental drivers considered for the cumulative effects assessment of global changes on the ecological communities of the Scotian Shelf."
  ) 
```

### Climate {#drclimate}

#### Sea surface temperature anomalies

The data used to characterize sea surface temperature anomalies come from the Atlantic Zone Monitorint Program [AZMP; @galbraith2021] from the Department of Fisheries and Oceans (DFO). A full description of the data and methods can be found in @galbraith2021, and we used a similar approach to that described by @beauchesne2020 to evaluate positive and negative temperature anomalies. Here, we provide a brief summary of the approach.

The surface layer is characterized using sea surface-water temperature (SST) monthly composites from Advanced Very High Resolution Radiometer (AVHRR) satellite images obtained from the National Oceanic and Atmospheric Administration (NOAA) and European Organization for the Exploitation of Meteorological Satellites (EUMETSAT). Images used are from DFO’s Maurice Lamontagne Institute at a 1 km resolution from 1985-2013 and from DFO’s Bedford Institute of Oceanography (BIO) Operational Remote Sensing group at a 1.5 km resolution since 2014. Monthly anomalies were constructed as the difference between monthly averages and the 1985-2010 climatological mean for each month.

We used temperature anomalies, *i.e.* deviations from long-term normal conditions between 1985 and 2010, to evaluate an annual index of stress associated with extreme temperatures between 2010 and 2021. Temperature anomalies were calculated using the difference between grid cell values with 1985-2010 climatological averages. Anomaly time series were normalized by their standard deviation (SD) to allow comparisons between areas with different temperature ranges. Grid cells whose monthly value exceeded ±0.5 standard deviation (SD) from the long-term average were considered as anomalous [@galbraith2021]. Outliers in the data were defined as those that fell beyond the interquartile range * 3, identified as extreme outliers by @tukey1977. Outlier values were capped to correspond to the 5th and 95th percentile values. Anomalies were divided into positive and negative anomalies and the absolute value of anomalies were used as the intensity of the surface-water temperature anomalies in a given cell. Only the months of May to November were included to avoid biases associated with the presence of ice cover. The sum of anomalies between 2010 and 2015, and between 2016 and 2021, were used to obtain a single value per cell per period considered for the assessment. 

#### Sea bottom temperature anomalies

TODO 

@dfo2022a

### Coastal {#drcoastal}

#### Coastal development 

We used lights at night as a proxy of the presence of coastal infrastructure development, as stable lights mostly capture human settlements and industrial sites. We used data from the Nighttime Lights Time Series. Nighttime light products are compiled by the Earth Observation Group at the National Oceanic and Atmospheric Administration's (NOAA) National Centers for Environmental Information (NCEI). They use globally available nighttime data obtained from the Visible Infrared Imaging Radiometer Suite (VIIRS) Day/Night Band (DNB) of the Defense Meteorological Satellite Program (DMSP) to characterize global average radiance ($nanoWatts$ $cm^{-2}$ $sr^{-1}$) composite images at a 15 arc-second (~200 m) resolution.v We used the annual Version 2 Nighttime VIIRS DNB composites between 2012 and 2021 [@elvidge2021] to characterize coastal development in coastal areas of the Scotiain Shelf. As the effects of coastal development are likely acute in its direct vicinity, we extracted average radiance values using a 2 km buffer around grid cells within 2 km of the coast. In each grid cell, we used the mean annual radiance as a proxy of the effects of coastal development in coastal areas. The first period (2010-2015) only contains 4 years of data, since radiance values are unavailable prior to 2012. 

#### Direct human impact

We used the sum of coastal populations as a proxy of direct human impact in coastal areas of the Scotian Shelf. We used Statistics Canada dissemination area and associated population counts from the 2016 [@statisticscanada2016a; @statisticscanada2016b; @statisticscanada2017] and 2021 [@statisticscanada2022; @statisticscanada2022a; @statisticscanada2022b] censuses to evaluated the population size in coastal areas. Dissemination areas are the smallest standard geographic area in which census data are disseminated. They combine to cover all of Canada and are highly variable in shapes and sizes. The census provides population count within the boundary of each dissemination area, which we used to evaluate total coastal population.

As the effects of direct human impacts are likely acute mostly in coastal areas
we calculated total population in grid cells within 2 km of the coast. Total
population was measured in a 10 km buffer around each coastal cell.
The total population in each buffer was the sum of intersecting dissemination
areas divided by the intersection area between buffers and dissemination areas:

$$DHI_j = \sum_{k=1}^{n_j} P_k * \frac{A_{j,k}}{A{tot, k}}$$

where $j$ is a buffered grid cell, $k$ is a dissemination area intersecting $j$,
$P$ is the population in $k$, $A$ is the area of the $k$ overlapping with $j$
and $A_{tot}$ is the total area of $k$. This approach was favoured to reduce the
effects of very large dissemination areas overlapping with buffers on a very
small percentage of their total area. We used the mean values from the 2016 and 2021 census for both periods of the assessment, as we did not have the data for the period between 2010 and 2015.

#### Inorganic pollution 


#### Nutrient input 


#### Organic pollution 


#### Population density 


### Fisheries {#drfisheries}

```{r fish}
path <- here::here(
  "data","data-integrated","fisheries_intensity-e2b7e6c4", 
  "fisheries_intensity-e2b7e6c4.yaml"
)
fish <- yaml::read_yaml(path)
nFishEvent <- sum(dplyr::bind_rows(fish$gear)$Freq)
nSpTarget <- length(fish$species$species_target) - 1
nSpCapt <- length(fish$species$species_capture)
```

The intensity and distribution of commercial fisheries on the Scotian Shelf was assessed using the Department of Fisheries and Oceans's logbook program [@dfo2021b]; these logbooks provide a thorough, if not exhaustive, overview of fisheries in eastern Canada. We used data between 2010 and 2021 to characterize the distribution and the intensity of commercial fisheries on the Scotian Shelf. This period was characterized by `r nFishEvent` fishing activities targeting `r nSpTarget` species and a total of `r nSpCapt` species were captured when considering bycatch. 

Fishing activities are performed using a variety of gear types (*e.g.* trap, trawl, dredge, driftnet, hand line, longline, scuba diving, purse seine, seine, beach seine and jig fishing), which can have different effects on species and habitats. Fisheries activities were therefore divided based on the type of environmental effects attributed to specific gear types according to the categories suggested by @halpern2008a and used by @beauchesne2020 for the Estuary and Gulf of St. Lawrence (Tables \@ref(tab:fishCat) and \@ref(tab:fishGear)):
demersal, destructive, high bycatch (DD)
demersal, non-destructive, high bycatch (DNH)
demersal, non-destructive, low bycatch (DNL)
pelagic, high bycatch (PHB)
pelagic, low bycatch (PLB)

Gear types can be further classified into fixed or mobile gear (Table \@ref(tab:fishGear)). We used the type of mobility to evaluate an expected area of effect for each fishing event, using radii of 200 and 2000 meters for fixed and mobile gear types, respectively [@beauchesne2020]. This approach, while decreasing precision, considers the potential uncertainty associated with fishing activity coordinates, gear mobility, and the absence of start and end coordinates for mobile gear. 

<br/>

```{r fishCat}
dplyr::bind_rows(fish$dataDescription) |>
  knitr::kable(
    col.names = c("Accronym","Category", "Description"),
    row.names = FALSE,
    caption = "Description of commercial fishing catagories used to characterize the intensity and distribution of fisheries on the Scotian Shelf."
  ) 
```

<br/>

```{r fishGear}
tab <- data.frame(
  gear = c("Trap","Bottom trawl","Drag","Gillnet","Line fishing",
           "Longline","Diving","Purse seine","Danish or Scottish seine",
           "Shoreline seine","Trap","Jigger"),
  class = c("DNH","DD","DD","PHB","PLB","PHB","DNL","PLB","DNH","DNH","DNH","PLB"),
  mob = c("Fixed","Mobile","Mobile","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed")
)

tab |>
  knitr::kable(
    col.names = c("Type of gear", "Category", "Mobility"),
    caption = "Classification of gear types based on their environmental effects and mobility. DD: demersal, destructive, high bycatch; DNH: demersal, non-destructive, high bycatch; DNL: demersal, non-destructive, low bycatch; PHB: pelagic, high bycatch; PLB: pelagic, low bycatch. Adapted from @beauchesne2020."
  )
```

<br/>


To characterize the intensity of fishing activities, we used a biomass yield density index by summing the total annual biomass captured in each grid cell covering the Scotian Shelf, yielding an assessment in kg per cell surface, *i.e.* $kg * km^{-2}$ in this case. For each period considered in the assessment (2010-2015 and 2016-2021) and each fishing category, we used the mean annual biomass captured per grid cell to capture the intensity and distribution of  as the assessment of the intensity and distribution of fisheries on the Scotian Shelf. 


### Marine traffic {#drmarinetraffic}

#### Shipping