---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      fig.path  = "man/figures/",
                      out.width = "100%")
```



# nceadfo


<!-- badges: start -->
[![License: GPL (>= 2)](https://img.shields.io/badge/License-GPL%20%28%3E%3D%202%29-blue.svg)](https://choosealicense.com/licenses/gpl-2.0/)
[![LifeCycle](https://img.shields.io/badge/lifecycle-experimental-orange)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Dependencies](https://img.shields.io/badge/dependencies-2/70-green?style=flat)](#)
<!-- badges: end -->


Research Compendium of the project **{{ PLEASE ADD A FEW WORDS }}**



### How to cite

Please cite this compendium as: 

> **{{ PLEASE ADD A CITATION }}**



### Content

This repository is structured as follow:

- [`data/`](https://github.com/Ecosystem-Assessments/compendium/tree/master/data):
contains all raw data required to perform analyses

- [`analyses/`](https://github.com/Ecosystem-Assessments/compendium/tree/master/analyses/):
contains R scripts to run each step of the workflow

- [`outputs/`](https://github.com/Ecosystem-Assessments/compendium/tree/master/outputs):
contains all the results created during the workflow

- [`figures/`](https://github.com/Ecosystem-Assessments/compendium/tree/master/figures):
contains all the figures created during the workflow

- [`R/`](https://github.com/Ecosystem-Assessments/compendium/tree/master/R):
contains R functions developed especially for this project and R scripts to run each step of the workflow

- [`man/`](https://github.com/Ecosystem-Assessments/compendium/tree/master/man):
contains help files of R functions

- [`DESCRIPTION`](https://github.com/Ecosystem-Assessments/compendium/tree/master/DESCRIPTION):
contains project metadata (author, date, dependencies, etc.)

- [`make.R`](https://github.com/Ecosystem-Assessments/compendium/tree/master/make.R):
main R script to run the entire project by calling each R script stored in the `analyses/` folder. NOTE: modify with a targets files



### Usage

- Clone this repository
- Open a terminal
- Build the Docker image with:

```sh
docker build -t "compendium" .
```

- Start a container based on this image:

```sh
docker run --rm -p 127.0.0.1:8787:8787 -e DISABLE_AUTH=true compendium
```

- On a web browser enter this URL: `127.0.0.1:8787`. A new RStudio Server
instance will be available.
- To run the analysis:

```{r eval = FALSE}
source("make.R")
```



### Notes

- All required packages, listed in the `DESCRIPTION` file, will be installed (if necessary)
- All required packages and R functions will be loaded
- Some analyses listed in the `make.R` might take time
