## Environmental drivers {#drmodule}

```{r drList}
dr <- read.csv(here::here("data","cea_modules","drivers_list.csv")) |>
      dplyr::select(-period) |>
      dplyr::distinct() 
nDr <- nrow(dr)
nCl <- sum(dr$group == "Climate")
nCo <- sum(dr$group == "Coastal")
nFs <- sum(dr$group == "Fisheries")
nMt <- sum(dr$group == "Marine traffic")
```

We characterized the distribution and intensity of `r nDr` environmental drivers on the Scotian Shelf during two distinct periods, *i.e.* 2010 to 2015 and 2016 to 2021. Temporal data was unavailable for certain drivers; in such instances, we used the same data for both periods (*e.g.* shipping). Drivers are divided in 4 groups: [climate](drclimate) (n = `r nCl`), [land-based / coastal](drcoastal) (n = `r nCo`), [fisheries](drfisheries) (n = `r nFs`) and [marine traffic](drmarinetraffic) (n = `r nMt`; Table \@ref(tab:drnm)). Drivers were log-transformed to avoid underestimating intermediate intensity values. All drivers were also scaled between 0 and 1 to obtain relative intensities and allow driver comparisons. Scaling was performed using the 99th quantile of intensity distribution as the upper bound to control for extreme values. For more details, see @beauchesne2020 and @halpern2019. The following subsections describe the approach used to evaluate the distribution and the intensity of each driver considered for the cumulative effects assessment. 

```{r drnm}
dplyr::select(dr, group, fullname, spatres, tempres, years, units, source) |>
dplyr::arrange(group, fullname) |>
  knitr::kable(
    col.names = c("Groups", "Drivers","Spatial resolution","Temporal resolution","Years","Units","Source"),
    row.names = FALSE,
    caption = "List of environmental drivers considered for the cumulative effects assessment of global changes on the ecological communities of the Scotian Shelf."
  ) 
```

### Climate {#drclimate}

#### Sea surface temperature anomalies

The data used to characterize sea surface temperature anomalies come from the Atlantic Zone Monitoring Program [AZMP; @galbraith2021] from the Department of Fisheries and Oceans (DFO). A full description of the data and methods can be found in @galbraith2021, and we used a similar approach to that described by @beauchesne2020 to evaluate positive and negative temperature anomalies. Here, we provide a brief summary of the approach.

The surface layer is characterized using sea surface-water temperature (SST) monthly composites from Advanced Very High Resolution Radiometer (AVHRR) satellite images obtained from the National Oceanic and Atmospheric Administration (NOAA) and European Organization for the Exploitation of Meteorological Satellites (EUMETSAT). Images used are from DFO’s Maurice Lamontagne Institute at a 1 km resolution from 1985-2013 and from DFO’s Bedford Institute of Oceanography (BIO) Operational Remote Sensing group at a 1.5 km resolution since 2014. Monthly anomalies were constructed as the difference between monthly averages and the 1985-2010 climatological mean for each month.

We used temperature anomalies, *i.e.* deviations from long-term normal conditions between 1985 and 2010, to evaluate an annual index of stress associated with extreme temperatures between 2010 and 2021. Temperature anomalies were calculated using the difference between grid cell values with 1985-2010 climatological averages. Anomaly time series were normalized by their standard deviation (SD) to allow comparisons between areas with different temperature ranges. Grid cells whose monthly value exceeded ±0.5 standard deviation (SD) from the long-term average were considered as anomalous [@galbraith2021]. Outliers in the data were defined as those that fell beyond the interquartile range * 3, identified as extreme outliers by @tukey1977. Outlier values were capped to correspond to the 5th and 95th percentile values. Anomalies were divided into positive and negative anomalies and the absolute value of anomalies were used as the intensity of the surface-water temperature anomalies in a given cell. Only the months of May to November were included to avoid biases associated with the presence of ice cover. The sum of anomalies between 2010 and 2015, and between 2016 and 2021, were used to obtain a single value per cell per period considered for the assessment. The anomalies were then resampled on the study grid used for this assessment.

#### Sea bottom temperature anomalies

The data used to characterize bottom-water temperature anomalies come from the Department of Fisheries and Oceans’ (DFO) temperature data from the summer groundfish survey on the Scotian Shelf. Bottom-water temperatures are interpolated from conductivity-temperature-depth (CTD) sampling performed annually at a 0.2 degrees resolution [@dfo2022a]. Anomalies are evaluated as a cell value for a given year minus that cell's value from a historical climatology. Anomalies were divided into positive and negative anomalies and the absolute value of anomalies were used as the intensity of the surface-water temperature anomalies in a given cell. Grid cells that exceeded 0.5 were considered as anomalous for this assessment. The mean anomaly intensity between 2010 and 2015, and between 2016 and 2019, for each grid cell was used to generate the final index of bottom-water temperature anomalies for each period considered.

### Coastal {#drcoastal}

#### Coastal development 

We used lights at night as a proxy of the presence of coastal infrastructure development, as stable lights mostly capture human settlements and industrial sites. We used data from the Nighttime Lights Time Series. Nighttime light products are compiled by the Earth Observation Group at the National Oceanic and Atmospheric Administration's (NOAA) National Centers for Environmental Information (NCEI). They use globally available nighttime data obtained from the Visible Infrared Imaging Radiometer Suite (VIIRS) Day/Night Band (DNB) of the Defense Meteorological Satellite Program (DMSP) to characterize global average radiance ($nanoWatts$ $cm^{-2}$ $sr^{-1}$) composite images at a 15 arc-second (~200 m) resolution.v We used the annual Version 2 Nighttime VIIRS DNB composites between 2012 and 2021 [@elvidge2021] to characterize coastal development in coastal areas of the Scotian Shelf. As the effects of coastal development are likely acute in its direct vicinity, we extracted average radiance values using a 2 km buffer around grid cells within 2 km of the coast. In each grid cell, we used the mean annual radiance as a proxy of the effects of coastal development in coastal areas. The first period (2010-2015) only contains 4 years of data, since radiance values are unavailable prior to 2012. 

#### Direct human impact

We used the sum of coastal populations as a proxy of direct human impact in coastal areas of the Scotian Shelf. We used Statistics Canada dissemination area and associated population counts from the 2016 [@statisticscanada2016a; @statisticscanada2016b; @statisticscanada2017] and 2021 [@statisticscanada2022; @statisticscanada2022a; @statisticscanada2022b] censuses to evaluated the population size in coastal areas. Dissemination areas are the smallest standard geographic area in which census data are disseminated. They combine to cover all of Canada and are highly variable in shapes and sizes. The census provides population count within the boundary of each dissemination area, which we used to evaluate total coastal population.

As the effects of direct human impacts are likely acute mostly in coastal areas we calculated total population in grid cells within 2 km of the coast. Total population was measured in a 10 km buffer around each coastal cell. The total population in each buffer was the sum of intersecting dissemination areas divided by the intersection area between buffers and dissemination areas:

$$DHI_j = \sum_{k=1}^{n_j} P_k * \frac{A_{j,k}}{A{tot, k}}$$

where $j$ is a buffered grid cell, $k$ is a dissemination area intersecting $j$, $P$ is the population in $k$, $A$ is the area of the $k$ overlapping with $j$ and $A_{tot}$ is the total area of $k$. This approach was favored to reduce the effects of very large dissemination areas overlapping with buffers on a very small percentage of their total area. We used the mean values from the 2016 and 2021 census for both periods of the assessment, as we did not have the data for the period between 2010 and 2015.

#### Land-based drivers

The data used to characterize the intensity of inorganic pollution, organic pollution, population density and nutrient input in coastal areas come from an atlas of land use for coastal watershed in the Maritimes Region, hereafter the Land Use Atlas [@guijarro-sabaniel2022]. We provide a brief summary of data and methods in this document. For more details, refer to @guijarro-sabaniel2022.

Coastal watershed boundaries in the Scotian Shelf Biogeographic Marine Management Sub-region were delineated using a digital elevation model (DEM) for each province was combined with national hydrographic data and calculating the flow direction and accumulation for each DEM. Stream orders for the stream network were then evaluated to locate the highest flow accumulation value for each watershed and given a value between 1 (lowest) and 4 (highest). The locations of these values were then used to identify pour points (*e.g.* river mouths) along the coast for each watershed. Spread of land stressors into coastal and marine environments was then modeled using a diffusive plume model using a density decay buffer based on stream order of the main pour points. 

The following sections describe how this model was used to characterize the intensity of inorganic pollution, nutrient input, organic pollution and population density in coastal areas of the Scotian Shelf.

##### Inorganic pollution

The intensity of inorganic pollution was modeled using the percent area covered by impervious surfaces in each watershed. Impervious surfaces refer to artificial structures such as roads and parking lots and industrial areas such as ports that are covered by water-resistant materials. Impervious surfaces are used as a proxy of inorganic pollution under the assumption that most of this pollution comes from urban runoff. Impervious surfaces were identified in watersheds using a combination of multiple datasets from the provinces New Brunswick and Nova Scotia. Refer to the *Land cover* section and Table 1 of the Material and Methods in @guijarro-sabaniel2022 for more details. 

##### Organic pollution 

We used the percent cover of agriculture land use in each watershed as a proxy of the intensity of organic pollution in each watershed. The percent area covered by agricultural land in each watershed was identified in watershed using a combination of multiple datasets from the provinces New Brunswick and Nova Scotia. Refer to the *Land cover* section and Table 1 of the Material and Methods in @guijarro-sabaniel2022 for more details. 

##### Population density 

This intensity of stress caused by watershed population in coastal areas was modeled using the population density in each watershed in $person * ha^{-1}$. Total population was estimated as the number of civic address datasets from New Brunswick and Nova Scotia multiplied by the average number of residents per household (*i.e.* 2.3) obtained from Statistics Canada. Population density was then evaluated as the total population in a watershed divided by the watershed area in hectares.


##### Nutrient input 

We used data from @kelly2021a to obtain the intensity of nutrient input in coastal areas of the Scotian Shelf. @kelly2021a used data from the Land Use Atlas to estimate nitrogen loading in coastal areas with the Nitrogen Loading Model (NLM) framework [@valiela1997; @valiela2000; @mciver2015; @nagel2018] to estimate total (dissolved) nitrogen from point and nonpoint source inputs in coastal areas from atmospheric deposition, fertilizer use, and wastewater disposal. Data and methods used to apply the NLM framework are described in detail in @kelly2021a.


### Fisheries {#drfisheries}

```{r fish}
path <- here::here(
  "data","data-integrated","fisheries_intensity-e2b7e6c4", 
  "fisheries_intensity-e2b7e6c4.yaml"
)
fish <- yaml::read_yaml(path)
nFishEvent <- sum(dplyr::bind_rows(fish$gear)$Freq)
nSpTarget <- length(fish$species$species_target) - 1
nSpCapt <- length(fish$species$species_capture)
```

The intensity and distribution of commercial fisheries on the Scotian Shelf was assessed using the Department of Fisheries and Oceans logbook program [@dfo2021b]; these logbooks provide a thorough, if not exhaustive, overview of fisheries in eastern Canada. We used data between 2010 and 2021 to characterize the distribution and the intensity of commercial fisheries on the Scotian Shelf. This period was characterized by `r nFishEvent` fishing activities targeting `r nSpTarget` species and a total of `r nSpCapt` species were captured when considering bycatch. 

Fishing activities are performed using a variety of gear types (*e.g.* trap, trawl, dredge, driftnet, hand line, longline, scuba diving, purse seine, seine, beach seine and jig fishing), which can have different effects on species and habitats. Fisheries activities were therefore divided based on the type of environmental effects attributed to specific gear types according to the categories suggested by @halpern2008a and used by @beauchesne2020 for the Estuary and Gulf of St. Lawrence (Tables \@ref(tab:fishCat) and \@ref(tab:fishGear)):
demersal, destructive, high bycatch (DD)
demersal, non-destructive, high bycatch (DNH)
demersal, non-destructive, low bycatch (DNL)
pelagic, high bycatch (PHB)
pelagic, low bycatch (PLB)

Gear types can be further classified into fixed or mobile gear (Table \@ref(tab:fishGear)). We used the type of mobility to evaluate an expected area of effect for each fishing event, using radii of 200 and 2000 meters for fixed and mobile gear types, respectively [@beauchesne2020]. This approach, while decreasing precision, considers the potential uncertainty associated with fishing activity coordinates, gear mobility, and the absence of start and end coordinates for mobile gear. 

<br/>

```{r fishCat}
dplyr::bind_rows(fish$dataDescription) |>
  knitr::kable(
    col.names = c("Accronym","Category", "Description"),
    row.names = FALSE,
    caption = "Description of commercial fishing catagories used to characterize the intensity and distribution of fisheries on the Scotian Shelf."
  ) 
```

<br/>

```{r fishGear}
tab <- data.frame(
  gear = c("Trap","Bottom trawl","Drag","Gillnet","Line fishing",
           "Longline","Diving","Purse seine","Danish or Scottish seine",
           "Shoreline seine","Trap","Jigger"),
  class = c("DNH","DD","DD","PHB","PLB","PHB","DNL","PLB","DNH","DNH","DNH","PLB"),
  mob = c("Fixed","Mobile","Mobile","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed","Fixed")
)

tab |>
  knitr::kable(
    col.names = c("Type of gear", "Category", "Mobility"),
    caption = "Classification of gear types based on their environmental effects and mobility. DD: demersal, destructive, high bycatch; DNH: demersal, non-destructive, high bycatch; DNL: demersal, non-destructive, low bycatch; PHB: pelagic, high bycatch; PLB: pelagic, low bycatch. Adapted from @beauchesne2020."
  )
```

<br/>


To characterize the intensity of fishing activities, we used a biomass yield density index by summing the total annual biomass captured in each grid cell covering the Scotian Shelf, yielding an assessment in kg per cell surface, *i.e.* $kg * km^{-2}$ in this case. For each period considered in the assessment (2010-2015 and 2016-2021) and each fishing category, we used the mean annual biomass captured per grid cell to capture the intensity and distribution of  as the assessment of the intensity and distribution of fisheries on the Scotian Shelf. 


### Marine traffic {#drmarinetraffic}

#### Shipping

The data used to characterize shipping come from the [Global Fishing Watch](https://globalfishingwatch.org/), which provides shipping data from Automatic Identification System (AIS) data globally. We obtained data between 2017 and 2020 that characterize monthly commercial shipping (*e.g.* cargo, reefer, tanker, bunker) as rasters with a 0.1 degree resolution, either as the number of vessels or the total hours of vessel presence in each grid cell [@gfw2022]. The data was also available either as direct observations from the AIS data, or interpolated to a regular interval of five minutes between points. We used the sum of the monthly interpolated number of all vessel types in each grid cell to obtain an assessment of annual shipping intensity on the Scotian Shelf. We then resampled the 0.1 degree resolution grid to our own study grid. We used the mean annual shipping intensity value between 2017 and 2020 for both periods considered in the assessment. 

