# Results {#results}

The following sections present the main results of the cumulative effects assessment. We begin by presenting the combined distributions of [taxa](#sprichres) and [environmental drivers](#cumdrres) as an introduction of the assessment results. These results essentially concatenate the areas of potential cumulative effects, since effects are constrained by the cooccurrence of taxa and environmental drivers. We then discuss the [spatial distribution](#ceares) and the [change in spatial distrbution](#diffres) between the periods of 2010-2015 and 2016-2021 for the species-scale and network-scale cumulative effects in the study area. Finally, we present results specific to the network-scale cumulative effects assessment, *i.e.* the [metanetwork of pathways of effects](#metares) and the [contribution of environmental drivers to direct and indirect effects](#contributionres) predicted by the cumulative effects assessment.

```{r urls}
beg <- "https://github.com/Ecosystem-Assessments/nceadfo/blob/main/figures/atlas/"
fin <- ".png?raw=true"
urls <- list()

# List of links to figures
urls$cumdrivers <- glue::glue("{beg}footprint/cumulative_drivers{fin}") 
urls$sprich <- glue::glue("{beg}footprint/species_richness{fin}") 
urls$ceasp <- glue::glue("{beg}cea/cea_species{fin}") 
urls$ceanet <- glue::glue("{beg}cea/cea_network{fin}") 
urls$ceadf <- glue::glue("{beg}difference/cea_difference{fin}") 
urls$meta <- glue::glue("{beg}metanetwork/metanetwork-2016_2021{fin}") 
urls$cont <- glue::glue("{beg}contribution/contribution_group-2016_2021{fin}") 

# List of captions to figures
spList <- read.csv(here::here("data","cea_modules","species_list.csv"))
nGr <- length(unique(spList$gr2))
cap <- list()
cap$cumdrivers <- paste("Distribution of cumulative drivers on the Scotian Shelf Bioregion in eastern Canada for the periods of 2010-2015 and 2015-2016. Cumulative drivers are evaluated as the sum of the normalized intensity (scaled between 0 and 1) of the", nDr, "drivers considered for the assessment ($E_{S_x} = \\sum_{j=1}^m S_{j,x}$); it identifies areas that are most exposed to the potential effects of the environmental drivers considered for the cumulative effects assessment. Refer to @beauchesne2020 for more details.")
cap$sprich <- paste("Distribution of species richness, as evaluated from the", nTx, "taxa considered in the cumulative effects assessment and the results of the species distribution modeling. See section \\@ref(spmodule) for more information.")
cap$ceasp <- paste("Species-scale cumulative effects assessment of", nDr, "drivers on", nTx, "taxa of the Scotian Shelf Bioregion in eastern Canada, using the @halpern2008a method. See section \\@ref(speciesmethod) for more details on the approach.")
cap$ceanet <- paste("Network-scale cumulative effects assessment of", nDr, "drivers on", nTx, "taxa of the Scotian Shelf Bioregion in eastern Canada, using the @beauchesne2020c method explicitly considering the underlying structure of the ecological community. See section \\@ref(networkmethod) for more details on the approach.")
cap$ceadf <- "Change in cumulative effects between the periods of 2010-2015 and 2016-2021 on the Scotian Shelf Bioregion in eastern Canada for the species-scale and network-scale assessments."
cap$meta <- paste("Multiplex network of network-scale cumulative effects on the Scotian Shelf Bioregion presenting the metaweb of interactions between the", nTx, "taxa considered (grey connections), the presence of an effect of individual drivers on each taxa (colored connections), the overall cumulative effect on each taxon (taxa point size) and the mean effect of drivers on all taxa (driver point size)")
cap$cont <- paste("Contribution of drivers to network-scale cumulative effects of taxa classified by taxonomic groups. Mean contribution of", nDr, "drivers to the cumulative effects ($C / km^2$) of", nTx, "taxa on the Scotian Shelf Bioregion in eastern Canada classified into", nGr, "taxonomic groups. Taxonomic groups for invertebrates (grey box) and vertebrates are divided at the phyla and classes level, respectively. For each group, the mean contribution per species was used to evaluate each driver's contribution to overall cumulative effects; this can be divided into their direct and indirect contributions to the cumulative effects on each taxon. Direct effects are those attributable to the effect of a driver on a focal species, while indirect effects are the mean effects of drivers spreading through all 3-species interactions a taxon is involved in. The total contribution to cumulative effect is the sum of the direct and indirect contributions of each driver.")

# ---
cap <- lapply(cap, function(x) glue::glue("{x} ***Click on figure to view in seperate tab.***"))
```

## Species richness {#sprichres}

This section of the assessment presents an evaluation of species richness, *i.e.* the number of taxa in each cell of the study grid, from the `r nTx` taxa considered for this assessment. It is important to mention here that the binary distribution of each taxa (*i.e.* pesence-absence) is based on observations for marine mammals and sea birds, while it is based on species distribution modeling predictions for all other marine species considered (see section \@ref(spmodule)). Figure \@ref(fig:sprich) presents the assessment of species richness in the study area, and the distribution of all individual taxa may be viewed in the [Atlas](@#atlas) of the assessment in section \@ref(spatlas).

We can see that the study area can be divided in two main zones of species richess, *i.e.* the continental shelf, including the Bay of Fundy, and the continental slope. Species richness is medium to high throughout the continental shelf and the Bay of Fundy, while it is lower on the continental slope. The transition between the shelf and the slope is however characterized by a latitudinal band of higher richness. Most of the shelf shows high species richness, especially the northeast and southwest tips of Nova-Scotia. The eastern part of the Bay of Fundy also has a particularly high species richness. It is however important to mention that the data used to characterize the distribution of species in the study area characterizes the continental shelf more extensively than the continental slope. Similarly, the survey data used to characterize the distribution of marine species comes from offshore trawl surveys that do not sample coastal waters; this results in lower species richness in coastal waters that should be interpreted in terms of sampling effort and design rather than as a general decrease in species richness. 

```{r sprich, out.width = "70%", echo=FALSE, fig.align='center', fig.link=urls$sprich, fig.cap=cap$sprich}
knitr::include_graphics(urls$sprich)
```

<br/>


## Cumulative drivers {#cumdrres}

This section of the assessment presents an evaluation of cumulative drivers, *i.e.* the sum of the normalized intensity of the `r nDr`drivers considered in the assessment, as a means to identify sites that are most exposed to the potential effects of environmental drivers in the study area. It does not provide an effects assessment, since it only considers the drivers; still, it allows us to assess the environments that are most likely to be affected by the stressors in our study area [see @beauchesne2020 for more information on cumulative drivers]. Figure \@ref(fig:cumuldr) presents the assessment of cumulative drivers for both periods considered, and all individual drivers may be viewed in the [Atlas](#atlas) of the assessment in section \@ref(stratlas). Cumulative drivers can also be explored interactively through the [*eDrivers* application](https://david-beauchesne.shinyapps.io/edriversapp_dev/), which is presented in section \@ref(edrivers) of this assessment.

As with species richness, the study area has two main zones of cumulative exposure, *i.e.* the continental shelf, including the Bay of Fundy, and the continental slope. Drivers are ubiquitous on continental shelf, which as a whole is exposed to at least medium levels of environmental drivers; the continental slope, meanwhile, is exposed to very few drivers when compared to the continental shelf. It is however important to consider that of the `r nDr` drivers considered, only positive and negative sea surface temperature anomalies and shipping are present on the continental slope (refer to section \@ref(stratlas) of the assessment atlas). Meanwhile, all drivers are found on the continental shelf and in the Bay of Fundy; offshore areas of the southwest continental shelf and the eastern end of the Bay of Fundy are slighlty more exposed to multiple drivers. The coasts of Nova-Scotia and of the whole Bay of Fundy are exposed to environmental drivers; this is true especially close to coastal towns, with Saint-John in the Bay of Fundy exhibiting the greatest values of cumulative drivers in the study area. 


```{r cumuldr, out.width = "100%", echo=FALSE, fig.align='center', fig.link=urls$cumdrivers, fig.cap=cap$cumdrivers}
knitr::include_graphics(urls$cumdrivers)
```

<br/>

## Cumulative effects {#ceares}

> The methods used to perform the species-scale and network-scale cumulative effects assessments are represented in sections \@ref(speciesmethod) and \@ref(networkmethod), respectively.

The main results of this assessment combine all data [modules](#modules) to assess the cumulative effects of environmental drivers on the ecological communities of the Scotian Shelf Bioregion. We present and interpret the results on the spatial distribution of cumulative effects from the species-scale (Figure \@ref(fig:spceares)) and network-scale (Figure \@ref(fig:nceares)) assessments jointly, as they identify similar spatial trends in cumulative effects. This is unsurprising, as the distribution of both assessments is driven mainly by the cooccurrence of environmental drivers and species. Considering species interactions amplifies or dampens the relative intensity of cumulative effects at the community scale, *i.e.* when all species are considered simultaneously; however, it refines the assessment for certain species such as marine mammals for which direct effects cannot properly capture the effects of drivers such as climate change that propagate through the ecological network. It should also be noted that values from the species-scale and network-scale assessments cannot be directly compared, as they are both relative to the component each consider. Hence, a cell with a greater value in the network-scale assessment cannot be interpreted as a greater impact; they can only be compared in relative form. See sections \@ref(metares) and \@ref(contributionres) for more on the direct and indirect cumulative effects, and refer to @beauchesne2020c for more details on the network-scale assessment. 

The assessments of cumulative effects reveal a stark contrast between the continental shelf, including the Bay of Fundy, and the continental slope; this result is in accordance with the lower representation of species and environmental drivers that we previously discussed from the assessments of [species richness](#sprichres) and [cumulative drivers](#cumdrres). This result should therefore also be interpreted as an absence data characterizing environmental drivers in the study area and that it is unlikely that ecological communities are not under any type of environmental threats. Still, it remains undeniable that the ecological communities of the continental shelf are more at risk from the cumulative effets of environmental drivers. 

On the continental shelf, the whole Bay of Fundy as well as the tip of southeastern Nova-Scotia extending offshore towards the continental slope appears to be the region most at risk from cumulative effects. The regions northeast of Sable Island, as well as the coastal areas south of Cape Breton are also at greater risk from cumulative effects. Areas of greatest cumulative effects seem to be strongly associated with regions affected by a combination of climate-related drivers, commercial fisheries and shipping (see [Atlas](@stratlas)). Meanwhile, the middle of the the continental shelf, besides a region associated with demersal destructive fisheries, as well as the northeastern corner of the study area close to the Laurential Channel seem to be less at risk from cumulative effects. 


```{r spceares, out.width = "100%", echo=FALSE, fig.align='center', fig.link=urls$ceasp, fig.cap=cap$ceasp}
knitr::include_graphics(urls$ceasp)
```

<br/>

```{r nceares, out.width = "100%", echo=FALSE, fig.align='center', fig.link=urls$ceanet, fig.cap=cap$ceanet}
knitr::include_graphics(urls$ceanet)
```

<br/>


## Change in cumulative effects {#diffres}

```{r evaldiff, echo = FALSE}
outdif <- here::here("output","cea_difference")
spcea <- stars::read_stars(here::here(outdif,"cea_species_difference.tif"))[[1]] 
spceamn <- round(mean(spcea, na.rm = TRUE),2)
spceasd <- round(sd(spcea, na.rm = TRUE),2)
ncea <- stars::read_stars(here::here(outdif,"cea_network_difference.tif"))[[1]]
nceamn <- round(mean(ncea, na.rm = TRUE),2)
nceasd <- round(sd(ncea, na.rm = TRUE),2)
```

> ***Note***: driver standardization is currently done on a driver per period basis, *e.g.* demersal fisheries for 2010-2015 were transformed using data from that period only. This makes the chagne in cumulative effects results less reliable reliable than if standardization was made from the joint distribution of demersal fishries from 2010-2015 and 2016-2021. This should be changed accordingly so that we may be confident in the results for changes over time.

The cumulative effects assessments were performed on two periods for environmental drivers when data permitted, the first spanning 2010 to 2015 and 2016 to 2021 (see section \@ref(drmodule) for more details). We thus evaluated the change in cumulative effects for the species-scale and the network-scale assessments (Figure \@ref(fig:ceadiff)). As with the cumulative effects themselves, the spatial distribution of relative change in cumulative effects is similar between both assessments. In general, there is a slight decrease in cumulative effects across the Scotiain Shelf Bioregion (species-scale assessment: `r spceamn` $\pm$ `r spceasd`; network-scale assessment: `r nceamn` $\pm$ `r nceasd`). Areas identified at higher risk of cumulative effects previously (see Section \@ref(ceares)) generally show a decrease in relative intensity, except for part of the Bay of Fundy, where a slight increase is predicted by both assessments. 


```{r ceadiff, out.width = "100%", echo=FALSE, fig.align='center', fig.link=urls$ceadf, fig.cap=cap$ceadf}
knitr::include_graphics(urls$ceadf)
```

<br/>

## Metanetwork {#metares}

Put metanetworks for direct and indirect effects as well.

```{r out.width = "90%", echo=FALSE, fig.align='center', fig.link=urls$meta, fig.cap=cap$meta}
knitr::include_graphics(urls$meta)
```

<br/>


## Driver contribution {#contributionres}

```{r out.width = "90%", echo=FALSE, fig.align='center', fig.link=urls$cont, fig.cap=cap$cont}
knitr::include_graphics(urls$cont)
```

<br/>
